Redshift Evolution of the Underlying Type Ia Supernova Stretch Distribution

### 2020-04-23 Nicolas.tex
Greg_corr_03 added

\abstract{
{The true nature of type Ia supernovae (SNe~Ia) remains largely unknown, and as survey statistics increase, the question of
astrophysical systematic uncertainties rises,  notably that of
the SN~Ia population redshift drift.}
{In this paper, we study the redshift evolution of SN~Ia
\textsc{\texttt{SALT2.4}} lightcurve stretch, a purely intrinsic SN property, to
probe this drift.}
{The SN stretch has been shown to strongly correlate with the SN's environment, notably with stellar age tracers. Following the evolution of
fraction of young and old SNe~Ia as predicted by \cite{rigault2018}, and
assuming constant underlying stretch
distribution for each population \mr{as a function of redshift}, model the expected SN~Ia stretch
distribution as a function of redshift. We test our prediction against data
from the literature chosen to have negligible selection effects to ensure
that any observed drift is indeed astrophysical and not observational.}
{We clearly demonstrate that the underlying SNe~Ia stretch distribution is
evolving as a function of redshift, and that the young/old drifting model is
a much better description of the data than any non-drifting model, including
the sample-based asymmetric distributions used by the Beams with Bias
Correction algorithm to correct Malmquist bias. Our favored underlying
stretch model is bimodal: a high-stretch mode shared by both young and old
environments, and a low-stretch mode exclusive to old environments.}
\nn{The
precise impact of the redshift evolution of the underlying SNe Ia population on
cosmology remains to be studied}. Yet, the astrophysical drift of the SN stretch
distribution does affect current Malmquist bias corrections and thereby
distances derived from SN affected by selection effects. We highlight that such
a bias will increase with surveys covering larger redshift ranges, which is
particularly important for LSST.}

@@

Type Ia supernovae (SNe Ia) are powerful cosmological distance indicators that
have enabled the discovery of the acceleration of the Universe's expansion
\citep{riess1998, perlmutter1999}. They remain today a key cosmological probe to
understand the properties of dark energy (DE) as it is the only tool able to
precisely map the recent expansion rate ($z<0.5$), when DE is driving it
\citep[e.g.][]{scolnicastro2020}. They also are key to directly measure the
Hubble Constant (H$_0$), provided one can calibrate their absolute magnitude
\citep{riess2016, freedman2019}. Interestingly, the value of H$_0$ derived when
the SNe~Ia are anchored on Cepheids \citep[the SH0ES
project,][]{riess2009,riess2016} is $\sim5\sigma$ higher than what is predicted
from cosmic microwave background (CMB) data measured by Planck assuming the
standard $\L$CDM \citep{planck2018,riess2019,reid2019}, or when the SN luminosity is anchored
at intermediate redshift by the baryon acoustic oscillation (BAO) scale \citep{feeney2019}. While using the tip of the
red giant branch technique in place of the Cepheids seem to favor an
intermediate value of H$_0$ \citep{freedman2019,freedman2020}, time delay
measurements from strong lensing seem to also favor high H$_0$ values
\citep{wong2019}.

The H$_0$ tension has received a lot of attention, as it could be a sign of new
fundamental physics. Yet, no simple solution is able to accommodate this H$_0$
tension when accounting for all other probes \citep{knox2019}, but see e.g. \cite{poulin2019}.
Alternatively, systematic effects affecting one or several of the aforementioned
analyses could also explain the tension. \cite{rigault2015} suggested that SNe
Ia from the Cepheid calibrator sample differ by construction from the Hubble
flow \nn{sample} ones as \nn{the former} strongly
favors young stellar population\nn{s}, while the latter not. This selection effect would impact the derivation of H$_0$ if
SNe~Ia from young and older environments differ in average standardized
magnitudes.

For the last decades, numerous analyses have studied the relation between SNe~Ia
and host properties, finding first that the standardized SNe~Ia magnitude
significantly depends on the host stellar mass, SNe~Ia from high-mass host being
brighter on average \cite[e.g.][]{kelly2010, sullivan2010, childress2013,
betoule2014, rigault2018, kim19}.  This mass-step correction is currently used
in cosmological analyses \citep[e.g.][]{betoule2014, scolnic2018a}, including
for deriving H$_0$ \citep{riess2016, riess2019}. Yet, the underlying connection
between the SNe and their host remains unclear when using global properties such
as the host stellar mass, which raises the question of the accuracy of the
correction. More recently, studies have used the local SN environment to probe
more direct connections between the SN and its environments \citep{rigault2013},
showing that local age tracers such as the Local specific Star Formation Rate
\nn{(LsSFR)} or the local color are more strongly correlated with the standardized
SN magnitude \citep{rigault2018, roman2018, kim18}, suggesting age as the
driving parameter underlying the mass-step. If true, this would have a
significant impact for cosmology, \mr{since the environmental correction to apply to SN standardization could strongly vary with redshift.} \citep{rigault2013, childress2014, scolnic2018a}. Yet, the
importance of local SN environmental studies remains highly debated
\cite[e.g.][]{jones2015, jones2019} and especially the impact \mr{of such an astrophysical bias} has on the derivation of
H$_0$ \citep{jones2015, riess2016, riess2018, rose2019}.

@@

In this paper, we take a step aside to probe the validity of our modeling of the
SN population, which we claim to be constituted of two age-populations
\citep{rigault2013,rigault2015,rigault2018}: one old and one younger, the former
having on average lower lightcurve stretches and being brighter after
standardization. We use the correlation between the SN age, as probed by the LsSFR, and the SN stretch to model
the expected evolution of the underlying SN stretch distribution as a function
of redshift. This modeling relies on three assumptions: (1) there are two
distinct populations of SNe~Ia; (2) the relative fraction of each of these
populations as a function of redshift follows the model presented in
\cite{rigault2018} and (3) the underlying distribution of stretch for each age
sample is constant. This paper aims at testing this specific model with datasets from the literature. \mr{Note that the progenitor age as traced by the LsSFR seems to capture physical feature intrinsic to the progenitor and/or explosion mechanism that the stretch alone is not capturing \citep{nordin2018}.}

We present in section~\ref{sec:sample} the sample we are using for this
analysis, derived from the Pantheon catalog \citep{scolnic2018a}. We discuss the
importance of obtaining a ``complete'' sample, i.e.\ representative of the true
underlying SNe Ia distribution, and how we build one from the Pantheon sample.
We then present in section~\ref{sec:modeling} our modeling of the distribution of stretch and our results are presented in
section~\ref{sec:results}. In this section we test whether the SN stretch distribution evolves as a function of redshift and if the aforementioned age model is in good agreement with this evolution. We briefly discuss these results in the context of SN cosmology in section~\ref{sec:bbc} and we conclude in
section~\ref{sec:ccl}.

@@

We base our analysis on the most recent comprehensive SNe~Ia
compilation, the Pantheon catalog from \cite{scolnic2018a}. A naive approach to
test the SN stretch redshift drift would be to simply compare the observed SN
stretch distributions in a few bins of redshift. In practice, however, \mr{differential} selection effects are affecting the observed SN stretch
distributions. Indeed, because the observed SN~Ia magnitude correlates with the
lightcurve stretch (and color), the first SNe~Ia that a magnitude-limited survey
will miss are the lowest-stretch (and reddest) ones. Consequently, if
magnitude-related selection effects are not accounted for, one might confuse
true population drift with survey properties, and conversely.

@@

% YC: Figs and Tables *after* their 1st ref. in the text!
\begin{figure}
    \centering
    \includegraphics[width=0.95\linewidth]{Article_figures/zmax_maglim_snls.pdf}
    \caption{\textsc{\texttt{SALT2.4}} stretch ($x_1$) and color ($c$)
        lightcurve parameters of SNe~Ia from the SDSS, PS1 and SNLS samples of
        the Pantheon catalog. The individual SNe are shown as blue dots. The ellipse $(x_1=\pm3, c=\pm0.3)$ is displayed, colored by
the corresponding standardized absolute magnitude using the $\alpha$ and $\beta$
coefficients from \cite{scolnic2018a}. The grey diagonal lines represent the
$(x_1, c)$ evolution for $m = m_{lim}$, for $z$ between $0.50$ and $z=1.70$
using SNLS's $m_{lim}$ of $24.8$~mag.}
    \label{fig:maglim}
\end{figure}

SNLS typically acquires SNe~Ia in the redshift range $0.4<z<0.8$; at these
redshifts, the rest-frame Bessel $B$ band roughly corresponds to the SNLS $i$
filter, that has a $5\sigma$ depth of
24.8~mag\footnote{\href{https://www.cfht.hawaii.edu/Science/CFHTLS/cfhtlsfinalreleaseexecsummary.html}{CFHT
final release website.}}. This converts to a $z_{lim}=0.60$, in agreement with
\cite{neill2006}, \cite{perrett2010} and \cite{bazin2011}.  Fig.~14 of
\cite[][\mr{see their section~5}]{perrett2010} suggests a lower limit of $z_{lim}=0.55$; both limits will be
considered, as discussed below.

Similarly, PS1 observes SNe~Ia in the range $0.2<z<0.4$, their $g$-band
$5\sigma$ depth is 23.1~mag \citep{rest2014}, which yields to $z_{lim}=0.30$ in
agreement with, e.g., Fig.~6 of \cite{scolnic2018a}. This figure is also
suggestive of a more conservative $z_{lim}=0.27$.

In a similar redshift range, SDSS has a limiting magnitude of 22.5
\citep{dilday2008,sako2008}, which would lead to a $z_{lim}=0.24$. However, the
SDSS surveys were more sensitive to limited spectroscopic resources;
\cite[][\mr{see their section~2}]{kessler2009} pointed out that during the first year of SDSS, SNe~Ia with
$r-mag<20.5$ were favored for spectroscopic follow-up, corresponding to a
redshift cut at $0.15$. For the rest of the SDSS survey, additional
spectroscopic resources were available, and \cite{kessler2009} and
\cite{dilday2008} show a \mr{reasonable} completeness up to
$z_{lim}=0.2$. Following these analyses, we will use $z_{lim}=0.2$ as the
baseline SDSS redshift limit.

The sample selection is summarized in Table~\ref{tab:sample}, and the redshift distribution of these three surveys is shown in Fig.~\ref{fig:cuts}.
As expected, the selected redshift limits roughly correspond to the peak of
these histograms. \mr{We show in section~\ref{sec:results} that this sample selection indeed provide a subset of SNe~Ia with insignificant selection effects when compared to state-of-the-art Malmquist correction techniques.}

\begin{table}
    \centering
    \caption{Composition of the SNe~Ia dataset used in this analysis.
    Conservative cuts are indicated in brackets.}
    \label{tab:sample}
    \begin{tabular}{l c c}
    \hline\hline%\\[-0.8em] % YC: don't, it's not your business to do that!
        Survey & $z_{lim}$ & $N_{\mathrm{SN}}$ \\%[0.15em]
        \hline%\\[-0.8em]
        SNf & -- & 114\\%[0.30em]
        SDSS & 0.20 (0.15) & 167 (82)\\%[0.30em]
        PS1 & 0.30 (0.27) & 160 (122)\\%[0.30em]        
        SNLS & 0.60 (0.55) & 102 (78)\\%[0.30em]
        HST & -- & 26\\%[0.30em]
        \hline%\\[-0.8em]
        Total & -- & 569 (422)\\%[0.30em]
        \hline
    \end{tabular}
\end{table}

\begin{figure}
    \centering
    \includegraphics[width=0.95\linewidth]{Article_figures/hist_surveys_cuts_55-cividis.pdf}
    \caption{\textit{From top to bottom}: Redshift histograms of SNe~Ia from
        the SDSS, PS1 and SNLS dataset respectively (data from Pantheon,
        \citealt{scolnic2018a}).  The colored parts represent the distribution
        of SNe~Ia kept in our analysis for they are supposedly free from
        selection bias (see section~\ref{sec:sample}). The darker (resp. lighter)
    color responds to the conservative (resp. fiducial) selection cut.}
    \label{fig:cuts}
\end{figure}

In addition, we use the SNe~Ia from the Nearby Supernova Factory
\citep[SNfactory,][]{aldering2002} published in \cite{rigault2018} and that have
been discovered from non-targeted searches (114 SNe~Ia, see their sections~3
and~4.2.2; \mr{SNe~Ia time series are published in \citealt{Saunders2020}, see also, Aldering in prep. and Ponder in prep.)}. The
spectroscopic follow-up was done over a redshift range of $0.02 < z <
0.09$~\citep[as in][]{rigault2018}, while the search was much deeper. As such,
these SNe~Ia are assumed to be a random sampling of the underlying SN
population. The SNfactory sample is particularly useful for studying SN property
drift, as it enables us to have a large SN~Ia sample at $z<0.1$.

@@

We present the stretch distribution and redshift histogram of these five surveys
up to their respective $z_{lim}$ in Fig.~\ref{fig:sample}. \mr{The evolution of the mean stretch is also shown in Fig.~\ref{fig:modelall}, where the data are split in redshift bins of regular sample size. We see that SNe~Ia at higher redshift have on average larger stretch ($0.34 \pm 0.10$ at $z\sim0.65$) than those at lower redshift ($-0.17\pm 0.10$ at $z\sim0.05$), suggesting that the underlying stretch distribution is drifting.}

\begin{figure}
    \centering
    \includegraphics[width=0.95\linewidth]{Article_figures/stretchs-cut_btw_hist_cumu_75-lb-cividis.pdf}
    \caption{\textit{Bottom:} \textsc{\texttt{SALT2.4}} lightcurve stretch as a
        function of redshift for each survey considered in this analysis (see
        legend). Solid (resp. open) markers correspond to the conservative
        (resp. fiducial) redshift cuts. \textit{Top:} combined redshift
    histogram in dark (resp. light) gray for the conservative (resp. fiducial)
redshift cuts.}
    \label{fig:sample}
\end{figure}

@@

\cite{rigault2018} presented a model for the evolution of the fraction of
younger and older SNe~Ia as a function of redshift following former work on
rates and delay time distributions \citep[e.g.,][]{mannucci2005,
scannapieco2005, sullivan2006, aubourg2008, childress2014, maozmannucci2014}.
In short, it was assumed that the number of ``young'' SNe~Ia follows the star
formation \nn{rate} (SFR) in the Universe, while the number of ``old'' SNe~Ia
follows the number of Gyr-old stars in the Universe, i.e. the stellar mass
(M$^*$). Hence, if we denote $\delta(z)$ (resp. $\psi(z) = 1-\delta(z)$) the
fraction of young (resp. old) SNe~Ia in the Universe as a function of redshift,
then the ratio $\delta/\psi$ is expected to follow the evolution of the specific
star formation rate (SFR/M$^*$), which goes as $(1+z)^{2.8}$ until $z\sim2$
\citep[e.g.,][]{tasca2015}. Since $\delta(0.05) \sim \psi(0.05)$
\citep{rigault2013,rigault2018,wiseman2020}, in agreement with rate expectations
\citep{mannucci2006,rodney2014}, \cite{rigault2018} concluded that
\begin{equation}
    \label{eq:delta}
    \delta(z) = \left( K^{-1} \times (1+z)^{-2.8} +1 \right)^{-1}
\end{equation}
with $K=0.87$. This model is comparable to the evolution predicted by
\cite{childress2014} based on SN rates in galaxies depending on their quenching
time as a function of their stellar mass.

@@

To model the evolution of the \nn{full} SN stretch distribution as a function of redshift, given our aforementioned model of the evolution of the fraction of
younger and older SNe~Ia with cosmic time, we need to model the SN stretch distribution for each age
subsample.

@@

\mr{The stretch probability distribution function (pdf) of a given SN will be the linear combination of the stretch distributions of these two population weighted by its probability $y^i$ to be young (see Section~\ref{sec:basemodelapplied}). But generally, the fraction of young SNIa as a function of redshift is given by $\delta$(z) (see Eq.~\ref{eq:delta}) and therefore, our redshift drift model of the underlying distribution of SNIa as a function of redshift $X_1(z)$ is given by}:
\begin{multline}\label{eq:stretchz}
    X_1(z) = \delta(z) \times \mathcal{N}(\mu_1,\sigma_1{}^2) + {} \\
    (1-\delta(z)) \times \left[ a\times\mathcal{N}(\mu_1,\sigma_1{}^2) +
    (1-a)\times\mathcal{N}(\mu_2,\sigma_2{}^2) \right].
\end{multline}

\subsection{Base model applied to data}
\label{sec:basemodelapplied}

Given the probability $y^i$ that a given SN is young, and assuming our Base (see section~\ref{sec:basemodel}), the probability to measure a \textsc{\texttt{SALT2.4}} stretch $x_1^i$ with an error d$x_1^i$ is given by:
\begin{align}\label{eq:likelihoodsnf}
    \prob{x^i_1}{\vec{\theta}; \mathrm{d}x^i_1, y^i} =
    y^i & \times
    \mathcal{N}\left(x^i_1 \mid \mu_1, \sigma_1{}^2+\mathrm{d}x^i_1{}^2\right) +
    \nonumber\\
    (1-y^i) &\times \bigg[
    a \times \mathcal{N}\left(x^i_1 \mid \mu_1,
    \sigma_1{}^2+\mathrm{d}x^i_1{}^2\right) +
    \nonumber\\
    & (1-a) \times \mathcal{N}\left(x^i_1 \mid \mu_2,
    \sigma_2{}^{2}+\mathrm{d}x^i_1{}^2\right) \bigg],
\end{align}

The maximum-likelihood estimate of the 5~free parameters
$\vec{\theta}\equiv({\mu_1,\mu_2,\sigma_1,\sigma_2,a})$ of the model is obtained
by minimizing the following:
\begin{equation}\label{eq:likelihood}
    -2\ln(L) = -2 \sum_i \ln \prob{x^{i}_{1}}{\vec{\theta};
    \mathrm{d}x^{i}_{1}, y^i}.
\end{equation}

@@

Results of fitting this model to all the 569 (resp. 422) SNe from the fiducial
(resp. conservative) sample are given Table~\ref{tab:modelresults}, and the
predicted redshift evolution of mean stretch (mean of Eq.~\ref{eq:stretchz})
illustrated as a blue band in Fig.~\ref{fig:modelall}. 
We see in this figure that the measured mean SN~Ia stretch per redshift bins of equal sample size closely follows our redshift drift modeling.
This is indeed what is expected if old environments favor low SN stretches
\citep[e.g.][]{howell2007} \emph{and} if the fraction of old SNe~Ia declines as
a function of redshift. See section~\ref{sec:results} for a more quantitative
discussion.

>>>>

\abstract{
The true nature of type Ia supernovae (SNe~Ia) remains largely unknown, and as
survey statistics increase, the question of astrophysical systematic
uncertainties rises, notably that of the SN~Ia population redshift drift. In
this paper, we study the redshift evolution of SN~Ia \textsc{\texttt{SALT2.4}}
lightcurve stretch, a purely intrinsic SN property, to probe this drift. The SN
stretch has been shown to strongly correlate with the SN's environment, notably
with stellar age tracers. Following the evolution of fraction of young and old
SNe~Ia as predicted by \cite{rigault2018}, and assuming constant underlying
stretch distribution for each population as a function of redshift, model the
expected SN~Ia stretch distribution as a function of redshift. We test our
prediction against data from the literature chosen to have negligible selection
effects to ensure that any observed drift is indeed astrophysical and not
observational. We clearly demonstrate that the underlying SNe~Ia stretch
distribution is evolving as a function of redshift, and that the young/old
drifting model is a much better description of the data than any non-drifting
model, including the sample-based asymmetric distributions used by the Beams
with Bias Correction algorithm to correct Malmquist bias. Our favored underlying
stretch model is bimodal: a high-stretch mode shared by both young and old
environments, and a low-stretch mode exclusive to old environments. The precise
impact of the redshift evolution of the underlying SNe Ia population on
cosmology remains to be studied. Yet, the astrophysical drift of the SN stretch
distribution does affect current Malmquist bias corrections and thereby
distances derived from SN affected by selection effects. We highlight that such
a bias will increase with surveys covering larger redshift ranges, which is
particularly important for LSST.
}

@@

Type Ia supernovae (SNe Ia) are powerful cosmological distance indicators that
have enabled the discovery of the acceleration of the Universe's expansion
\citep{riess1998, perlmutter1999}. They remain today a key cosmological probe to
understand the properties of dark energy (DE) as it is the only tool able to
precisely map the recent expansion rate ($z<0.5$), when DE is driving it
\citep[e.g.][]{scolnicastro2020}. They also are key to directly measure the
Hubble Constant (H$_0$), provided one can calibrate their absolute magnitude
\citep{riess2016, freedman2019}. Interestingly, the value of H$_0$ derived when
the SNe~Ia are anchored on Cepheids \citep[the SH0ES project,][]{riess2009,
riess2016} is $\sim5\sigma$ higher than what is predicted from cosmic microwave
background (CMB) data measured by Planck assuming the standard $\L$CDM
\citep{planck2018, riess2019, reid2019}, or when the SN luminosity is anchored
at intermediate redshift by the baryon acoustic oscillation (BAO) scale
\citep{feeney2019}. While using the tip of the red giant branch technique in
place of the Cepheids seem to favor an intermediate value of H$_0$
\citep{freedman2019, freedman2020}, time delay measurements from strong lensing
seem to also favor high H$_0$ values \citep{wong2019}.

The H$_0$ tension has received a lot of attention, as it could be a sign of new
fundamental physics. Yet, no simple solution is able to accommodate this H$_0$
tension when accounting for all other probes \citep{knox2019}, but see e.g.
\cite{poulin2019}. Alternatively, systematic effects affecting one or several
of the aforementioned analyses could also explain the tension.
\cite{rigault2015} suggested that SNe Ia from the Cepheid calibrator sample
differ by construction from the Hubble flow sample ones as the former
strongly favors young stellar populations, while the latter not. This
selection effect would impact the derivation of H$_0$ if SNe~Ia from young and
older environments differ in average standardized magnitudes.

For the last decades, numerous analyses have studied the relation between SNe~Ia
and host properties, finding first that the standardized SNe~Ia magnitude
significantly depends on the host stellar mass, SNe~Ia from high-mass host being
brighter on average \cite[e.g.][]{kelly2010, sullivan2010, childress2013,
betoule2014, rigault2018, kim19}. This mass-step correction is currently used in
cosmological analyses \citep[e.g.][]{betoule2014, scolnic2018a}, including for
deriving H$_0$ \citep{riess2016, riess2019}. Yet, the underlying connection
between the SNe and their host remains unclear when using global properties such
as the host stellar mass, which raises the question of the accuracy of the
correction. More recently, studies have used the local SN environment to probe
more direct connections between the SN and its environments \citep{rigault2013},
showing that local age tracers such as the Local specific Star Formation Rate
(LsSFR) or the local color are more strongly correlated with the standardized SN
magnitude \citep{rigault2018, roman2018, kim18}, suggesting age as the driving
parameter underlying the mass-step. If true, this would have a significant
impact for cosmology, since the environmental correction to apply to SN
standardization could strongly vary with redshift. \citep{rigault2013,
childress2014, scolnic2018a}. Yet, the importance of local SN environmental
studies remains highly debated \cite[e.g.][]{jones2015, jones2019} and
especially the impact of such an astrophysical bias has on the derivation of
H$_0$ \citep{jones2015, riess2016, riess2018, rose2019}.

@@

In this paper, we take a step aside to probe the validity of our modeling of the
SN population, which we claim to be constituted of two age-populations
\citep{rigault2013,rigault2015,rigault2018}: one old and one younger, the former
having on average lower lightcurve stretches and being brighter after
standardization. We use the correlation between the SN age, as probed by the
LsSFR, and the SN stretch to model the expected evolution of the underlying SN
stretch distribution as a function of redshift. This modeling relies on three
assumptions: (1) there are two distinct populations of SNe~Ia; (2) the relative
fraction of each of these populations as a function of redshift follows the
model presented in \cite{rigault2018} and (3) the underlying distribution of
stretch for each age sample is constant. This paper aims at testing this
specific model with datasets from the literature. Note that the progenitor age
as traced by the LsSFR seems to capture physical feature intrinsic to the
progenitor and/or explosion mechanism that the stretch alone is not capturing
\citep{nordin2018}.

We present in Section~\ref{sec:sample} the sample we are using for this
analysis, derived from the Pantheon catalog \citep{scolnic2018a}. We discuss the
importance of obtaining a ``complete'' sample, i.e.\ representative of the true
underlying SNe Ia distribution, and how we build one from the Pantheon sample.
We then present in Section~\ref{sec:modeling} our modeling of the distribution
of stretch and our results are presented in Section~\ref{sec:results}. In this
section we test whether the SN stretch distribution evolves as a function of
redshift and if the aforementioned age model is in good agreement with this
evolution. We briefly discuss these results in the context of SN cosmology in
Section~\ref{sec:discussion} and we conclude in Section~\ref{sec:ccl}.

@@

We base our analysis on the most recent comprehensive SNe~Ia compilation, the
Pantheon catalog from \cite{scolnic2018a}. A naive approach to test the SN
stretch redshift drift would be to simply compare the observed SN stretch
distributions in a few bins of redshift. In practice, however, differential
selection effects are affecting the observed SN stretch distributions. Indeed,
because the observed SN~Ia magnitude correlates with the lightcurve stretch (and
color), the first SNe~Ia that a magnitude-limited survey will miss are the
lowest-stretch (and reddest) ones. Consequently, if magnitude-related selection
effects are not accounted for, one might confuse true population drift with
survey properties, and conversely.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.95\linewidth]{Article_figures/zmax_maglim_snls.pdf}
    \caption{\textsc{\texttt{SALT2.4}} stretch ($x_1$) and color ($c$)
        lightcurve parameters of SNe~Ia from the SDSS, PS1 and SNLS samples of
        the Pantheon catalog. The individual SNe are shown as blue dots. The
        ellipse $(x_1=\pm3, c=\pm0.3)$ is displayed, colored by the
        corresponding standardized absolute magnitude using the $\alpha$ and
        $\beta$ coefficients from \cite{scolnic2018a}. The grey diagonal lines
        represent the $(x_1, c)$ evolution for $m = m_{lim}$, for $z$ between
        $0.50$ and $z=1.70$ using SNLS's $m_{lim}$ of $24.8$~mag.}
    \label{fig:maglim}
\end{figure}

SNLS typically acquired SNe~Ia in the redshift range $0.4<z<0.8$; at these
redshifts, the rest-frame Bessel $B$ band roughly corresponds to the SNLS $i$
filter, that has a $5\sigma$ depth of
24.8~mag\footnote{\href{https://www.cfht.hawaii.edu/Science/CFHTLS/cfhtlsfinalreleaseexecsummary.html}{CFHT
final release website.}}. This converts to a $z_{lim}=0.60$, in agreement with
\cite{neill2006}, \cite{perrett2010} and \cite{bazin2011}. Fig.~14 of
\cite[][see their Section~5]{perrett2010} suggests a lower limit of
$z_{lim}=0.55$; both limits will be considered, as discussed below.

Similarly, PS1 observed SNe~Ia in the range $0.2<z<0.4$, their $g$-band
$5\sigma$ depth is 23.1~mag \citep{rest2014}, which yields to $z_{lim}=0.30$ in
agreement with, e.g., Fig.~6 of \cite{scolnic2018a}. This figure is also
suggestive of a more conservative $z_{lim}=0.27$.

In a similar redshift range, SDSS has a limiting magnitude of 22.5
\citep{dilday2008, sako2008}, which would lead to a $z_{lim}=0.24$. However, the
SDSS surveys were more sensitive to limited spectroscopic resources; \cite[][see
their Section~2]{kessler2009} pointed out that during the first year of SDSS,
SNe~Ia with $r-mag<20.5$ were favored for spectroscopic follow-up, corresponding
to a redshift cut at $0.15$. For the rest of the SDSS survey, additional
spectroscopic resources were available, and \cite{kessler2009} and
\cite{dilday2008} show a reasonable completeness up to $z_{lim}=0.2$. Following
these analyses, we will use $z_{lim}=0.2$ as the baseline SDSS redshift limit.

The sample selection is summarized in Table~\ref{tab:sample}, and the redshift
distribution of these three surveys is shown in Fig.~\ref{fig:cuts}. As
expected, the selected redshift limits roughly correspond to the peak of these
histograms. We show in Section~\ref{sec:results} that this sample selection
indeed provides a subset of SNe~Ia with insignificant selection effects when
compared to state-of-the-art Malmquist correction techniques.

\begin{table}
    \centering
    \caption{Composition of the SNe~Ia dataset used in this analysis.
    Conservative cuts are indicated in brackets.}
    \label{tab:sample}
    \begin{tabular}{l c c}
        \hline\hline
        Survey & $z_{lim}$ & $N_{\mathrm{SN}}$ \\
        \hline
        SNf   & --          & 114 \\
        SDSS  & 0.20 (0.15) & 167 (82)\\
        PS1   & 0.30 (0.27) & 160 (122)\\
        SNLS  & 0.60 (0.55) & 102 (78)\\
        HST   & --          & 26 \\
        \hline
        Total & --          & 569 (422) \\
        \hline
    \end{tabular}
\end{table}

\begin{figure}
    \centering
    \includegraphics[width=0.95\linewidth]{Article_figures/hist_surveys_cuts_55-cividis.pdf}
    \caption{\textit{From top to bottom}: Redshift histograms of SNe~Ia from
        the SDSS, PS1 and SNLS dataset respectively (data from Pantheon,
        \citealt{scolnic2018a}). The colored parts represent the distribution
        of SNe~Ia kept in our analysis for they are supposedly free from
        selection bias (see Section~\ref{sec:sample}). The darker (resp. lighter)
        color responds to the conservative (resp. fiducial) selection cut.}
    \label{fig:cuts}
\end{figure}

In addition, we use the SNe~Ia from the Nearby Supernova Factory
\citep[SNfactory,][]{aldering2002} published in \cite{rigault2018} and that have
been discovered from non-targeted searches (114 SNe~Ia, see their sections~3
and~4.2.2; SNe~Ia time series are published in \citealt{Saunders2020}
\nn{\textit{Greg: for now you may need to cite SNEMO and SUGAR to remove this
``?''}}, see also Aldering in prep. and Ponder in prep.). The spectroscopic
follow-up was done over a redshift range of $0.02 < z < 0.09$~\citep[as
in][]{rigault2018}, while the search was much deeper. As such, these SNe~Ia are
assumed to be a random sampling of the underlying SN population. The SNfactory
sample is particularly useful for studying SN property drift, as it enables us
to have a large SN~Ia sample at $z<0.1$.

@@

We present the stretch distribution and redshift histogram of these five surveys
up to their respective $z_{lim}$ in Fig.~\ref{fig:sample}. The evolution of
the mean stretch is also shown in Fig.~\ref{fig:modelall}, where the data are
split in redshift bins of regular sample size. We see that SNe~Ia at higher
redshift have on average larger stretch ($0.34 \pm 0.10$ at $z\sim0.65$) than
those at lower redshift ($-0.17\pm 0.10$ at $z\sim0.05$), suggesting that the
underlying stretch distribution is drifting.

\begin{figure}
    \centering
    \includegraphics[width=0.95\linewidth]{Article_figures/stretchs-cut_btw_hist_cumu_75-lb-cividis.pdf}

    \caption{\textit{Bottom:} \textsc{\texttt{SALT2.4}} lightcurve stretch as a
        function of redshift for each survey considered in this analysis (see
        legend). Solid (resp.\ open) markers correspond to the conservative
        (resp.\ fiducial) redshift cuts. \textit{Top:} combined redshift
        histogram in dark (resp.\ light) gray for the conservative (resp.
        \ fiducial) redshift cuts.}

    \label{fig:sample}
\end{figure}

@@

\cite{rigault2018} presented a model for the evolution of the fraction of
younger and older SNe~Ia as a function of redshift following former work on
rates and delay time distributions \citep[e.g.,][]{mannucci2005,
scannapieco2005, sullivan2006, aubourg2008, childress2014, maozmannucci2014}.
In short, it was assumed that the number of ``young'' SNe~Ia follows the star
formation rate (SFR) in the Universe, while the number of ``old'' SNe~Ia
follows the number of Gyr-old stars in the Universe, i.e. the stellar mass
(M$^*$). Hence, if we denote $\delta(z)$ (resp. $\psi(z) = 1-\delta(z)$) the
fraction of young (resp. old) SNe~Ia in the Universe as a function of redshift,
then the ratio $\delta/\psi$ is expected to follow the evolution of the specific
star formation rate (SFR/M$^*$), which goes as $(1+z)^{2.8}$ until $z\sim2$
\citep[e.g.,][]{tasca2015}. Since $\delta(0.05) \sim \psi(0.05)$
\citep{rigault2013,rigault2018,wiseman2020}, in agreement with rate expectations
\citep{mannucci2006,rodney2014}, \cite{rigault2018} concluded that
\begin{equation}
    \label{eq:delta}
    \delta(z) = \left( K^{-1} \times (1+z)^{-2.8} +1 \right)^{-1}
\end{equation}
with $K=0.87$. This model is comparable to the evolution predicted by
\cite{childress2014} based on SN rates in galaxies depending on their quenching
time as a function of their stellar mass.

@@

To model the evolution of the full SN stretch distribution as a function of
redshift, given our aforementioned model of the evolution of the fraction of
younger and older SNe~Ia with cosmic time, we need to model the SN stretch
distribution for each age subsample.

@@

The stretch probability distribution function (pdf) of a given SN will be the
linear combination of the stretch distributions of these two population weighted
by its probability $y^i$ to be young (see Section~\ref{sec:basemodelapplied}).
But generally, the fraction of young SNe~Ia as a function of redshift is given
by $\delta$(z) (see Eq.~\ref{eq:delta}) and therefore, our redshift drift model
of the underlying distribution of SNe~Ia as a function of redshift $X_1(z)$ is
given by:
\begin{multline}\label{eq:stretchz}
    X_1(z) = \delta(z) \times \mathcal{N}(\mu_1,\sigma_1{}^2) + {} \\
    (1-\delta(z)) \times \left[ a\times\mathcal{N}(\mu_1,\sigma_1{}^2) +
    (1-a)\times\mathcal{N}(\mu_2,\sigma_2{}^2) \right].
\end{multline}

\subsection{Base model applied to data}\label{sec:basemodelapplied}

Given the probability $y^i$ that a given SN is young, and assuming our Base (see
Section~\ref{sec:basemodel}), the probability to measure a
\textsc{\texttt{SALT2.4}} stretch $x_1^i$ with an error d$x_1^i$ is given by:
\begin{align}\label{eq:likelihoodsnf}
    \prob{x^i_1}{\vec{\theta}; \mathrm{d}x^i_1, y^i} =
    y^i & \times
    \mathcal{N}\left(x^i_1 \mid \mu_1, \sigma_1{}^2+\mathrm{d}x^i_1{}^2\right) +
    \nonumber\\
    (1-y^i) &\times \bigg[
    a \times \mathcal{N}\left(x^i_1 \mid \mu_1,
    \sigma_1{}^2+\mathrm{d}x^i_1{}^2\right) +
    \nonumber\\
    & (1-a) \times \mathcal{N}\left(x^i_1 \mid \mu_2,
    \sigma_2{}^{2}+\mathrm{d}x^i_1{}^2\right) \bigg]
\end{align}

The maximum-likelihood estimate of the 5~free parameters
$\vec{\theta}\equiv({\mu_1,\mu_2,\sigma_1,\sigma_2,a})$ of the model is obtained
by minimizing the following:
\begin{equation}\label{eq:likelihood}
    -2\ln(L) = -2 \sum_i \ln \prob{x_1^i}{\vec{\theta};
    \mathrm{d}x_1^i, y^i}.
\end{equation}

@@

Results of fitting this model to all the 569 (resp. 422) SNe from the fiducial
(resp. conservative) sample are given Table~\ref{tab:modelresults}, and the
predicted redshift evolution of mean stretch (mean of Eq.~\ref{eq:stretchz})
illustrated as a blue band in Fig.~\ref{fig:modelall}. We see in this figure
that the measured mean SN~Ia stretch per redshift bins of equal sample size
closely follows our redshift drift modeling. This is indeed what is expected if
old environments favor low SN stretches \citep[e.g.][]{howell2007} \emph{and} if
the fraction of old SNe~Ia declines as a function of redshift. See
Section~\ref{sec:results} for a more quantitative discussion.
